\\Large The heat is on: Rising temperatures alter how and where mammals move

### 2025-09-02 writing/manuscript.Rmd
reviewed methods

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behavior cycles. Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023. We show that temperature is a strong determinant of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century. We found that species did not respond uniformly to rising temperatures, so community-level responses will likely be complex as some species move more and others less. Median habitat selection strength decreased for all species, but some species showed increased selection strength for higher altitudes. Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effectively and proactively understanding mammalian movement behavior and protecting high-quality habitat. We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

* @gerlich_species_2025: effects of temperature on fly movement in the arctic
* "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C" (https://www.sciencedirect.com/science/article/pii/S2214581823000770#fig0015)
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
* changes in temperature change phelogy and movement behavior as a consequence @gerlich_species_2025
* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
\clearpage

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016]. By altering mammals' movement speed and habitat use [jointly, their movement behavior: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behavior [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. Behavioral changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more rapid and serious physiological damage [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, particularly if they have a large body size [@dyer_travel_2023; @verzuh_behavioural_2023]. Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with growing anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019]. However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behavior and quantitative predictions of how their behavior will change in the current century.

@@

This paper provides a quantifyies the effects of proximal air temperature on mammalian movement behavior, namely species' movement rates and habitat selection. We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement behavior during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We discuss the consequences of these changes at the individual, population, and community levels. We also provide suggestions for future research, including sampling design to overcome some common data limitations.

@@

knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with resolved speed (Has speed) for each of the species in this study.', escape = FALSE, longtable = FALSE, booktabs = TRUE)
```

@@

This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig. \@ref(fig:bc-map)); (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017]. While we acknowledge that forest type, structure, and age are important variables, we do not include such variables in this study for the sake of parsimony and comparability across species. We detail the data sources and analyses in the sections below.

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public British Columbia Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour (Appendix A).

### Historical temperature data and temperature projections

\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].

We obtained rasters of projected monthly average temperature for the study region at a 0.08$^\circ$ spatial resolution from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 reanalized data for western Canada from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.

```{r, eval=FALSE, echo=FALSE}
m_gammals <- gam(list(
  # linear predictor for the mean parameter
  monthly_var ~
    s(monthly_mean, bs = 'cr', k = 10) +
    s(month, bs = 'cc', k = 10) + # using a cyclic cubic spline basis
    s(long, lat, bs = 'ds', k = 500) +
    ti(long, lat, month, bs = c('ds', 'cc'), d = c(2, 1), k = c(100, 10)),
  # linear predictor for the scale parameter
  ~ s(monthly_mean, bs = 'cr', k = 10) +
    s(month, bs = 'cc', k = 10) +
    s(long, lat, bs = 'ds', k = 100)),
  family = gammals(),
  data = d,
  method = 'REML',
  knots = list(month = c(0.5, 12.5)))
```

We simulated hourly variation in future weather by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS. For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile.

@@

\noindent We estimated percent forest cover and distance from water using the rasters created by @tuanmu_global_2014. We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needle-leaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively). We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package. Finally, we obtained two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [version 0.99.0, @hollister_elevatr_2023]: We used a raster with a zoom of 6 (0.009$^\circ$) for model fitting and one with a zoom of 3 (0.08$^\circ$) for downloading climate change projections via `climatenaR`. All final rasters and code are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

@@

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The 18 animals with insufficiently fine sampling to support movement models with finite speed [@noonan_scale-insensitive_2019] were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).

Since `ctmm``s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using a $k$-means algorithm with 2 clusters for each species' detrended speeds. When the algorithm clearly failed to discriminate between states, we estimated the split point using the inflection points in histograms of the detrended speeds (Fig. B1).

## Estimating the effects of temperature on mammals' movement

\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models.

### Disentangling direct and indirect effects

\noindent In this study, we separate the effects of temperature on mammalian movement into direct and indirect effects. We call "direct" all effects that impact movement behavior causally without first impacting another variable (Fig. \@ref(fig:dags)). For example, in Fig. \@ref(fig:dags)A, whether a grizzly bear is moving in a given moment directly depends on time of day and day of year, since it will not be moving when sleeping or hibernating. However, time of day and day of year also impact its movement behavior indirectly because its sensitivity to temperature will also depend on time of day and day of year. Thus, both time of day and day of year also impact movement frequency indirectly by altering the effect of temperature on movement frequency. Failing to disentangle the effects of these three variables will lead to incorrect and likely confusing estimates for the causal effect of temperature on movement behavior, since causal and non-causal relationships would be indistinguishable [@mcelreath_statistical_2020].

Indirect effects of temperature also include all effects that can be controlled by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022]. For example, in Fig. \@ref(fig:dags)B, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can mitigate these indirect effects of temperature by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022]. In this case, it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly).

>>>>

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behaviour cycles. Consequently, it is still unclear how mammals will adapt their movement behaviour in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023. We show that temperature is a strong determinant of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century. We found that species did not respond uniformly to rising temperatures, so community-level responses will likely be complex as some species move more and others less. Median habitat selection strength decreased for all species, but some species showed increased selection strength for higher altitudes. Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behaviour and community structure will become crucial for effectively and proactively understanding mammalian movement behaviour and protecting high-quality habitat. We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

* @gerlich_species_2025: effects of temperature on fly movement in the arctic
* "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C" (https://www.sciencedirect.com/science/article/pii/S2214581823000770#fig0015)
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
* changes in temperature change phelogy and movement behaviour as a consequence @gerlich_species_2025
* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
\clearpage

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016]. By altering mammals' movement speed and habitat use [jointly, their movement behaviour: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behaviour [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. Behavioural changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size [@dyer_travel_2023; @verzuh_behavioural_2023]. Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last century, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020], which will be exacerbated by more frequent and intense extreme events [@bunnell_global_2011] and anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019]. However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of how their behaviour will respond to climate change in the current century [@ipcc_climate_2023].

@@

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement behaviour during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We discuss the consequences of these changes in movement behaviour at the individual, population, and community levels. We also provide directions for future research and solutions for overcoming some common data limitations.

@@

knitr::kable(dt_data, format = 'markdown', caption = "Summary statistics of each species' set of GPS data, including: the start and end of each set of GPS telemetries, the median sampling interval ($\\Delta t$; stratified by animal), number of animals (Animals), and number of animals with finite speed estimates (Has speed).", escape = FALSE, longtable = FALSE, booktabs = TRUE)
```

@@

This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig. \@ref(fig:bc-map)); (2) historical hourly air temperature from the ERA5 reanalyzed dataset [European Center for Medium-range Weather Forecasting$;$ @hersbach_era5_2023]; (3) temporally static rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs$;$ @riahi_shared_2017]. While we acknowledge there are other important variables (e.g., forest type, structure, and age; terrain slope and aspect), we limited our models to only use three simple variables for the sake of parsimony and comparability across species. See the discussion section for additional detail on the advantages and limitations of using such simple models.

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022]. Boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018]. ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, we removed a total of 2396 problematic GPS locations (0.16%, including collar calibration data) after inspecting diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated relatively stationary behaviour (Appendix A).

### Historical temperature data and climate change projections

\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal air temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [v. 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].

We used the the `climatenaR` package [v. 1.0, @burnett_climatenar_2023] for `R` and ClimateNA [v. 7.4.2$;$ @wang_locally_2016; @mahony_global_2022] to obtain projected average monthly temperatures from 2020 to 2100 at a 0.08$^\circ$ spatial resolution. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the historical data for western Canada from 1998 to 2023 (extremes included) to estimate the monthly variation in temperature, which we defined as $\sigma^2_T$, the variance at location $\langle x, y \rangle$ in month $m$ for each year (e.g., the variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005). We then modeled the estimated variance using a Generalized Additive Model for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] using the `mgcv` package for `R` [v. 1.9-1, @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of: each year's estimated within-pixel monthly mean temperature ($\hat \mu_T$), month ($m$), a two-dimensional smooth of spatial coordinates $\langle x, y \rangle$, and a tensor product interaction term of month and space to allow for seasonal terms to vary smoothly over space. The linear predictor for the scale term, which governs the mean-variance relationship [table 3.1 on p. 104 of @wood_generalized_2017], included smooth terms of the monthly mean temperature, month, and space. We did not include smooth terms of year to avoid unrealistic projections when extrapolating beyond past 2023. The model was thus

\begin{equation}\label{eq:temp-var-model}
\begin{cases}
\sigma^2_T \sim \text{Gamma}(\mu_{\sigma^2_T}, \nu_{\sigma^2_T}) \\
\mathbb E(\sigma^2_T) = \mu_{\sigma^2_T} \\
\mathbb V(\sigma^2_T) = (\mu_{\sigma^2_T})^2 / \nu_{\sigma^2_T} \\
\log(\mu_{\sigma^2}) = \beta_{L,0} + f_{L,1}(\mu_T) + f_{L,2}(m) + f_{L,3}(x, y) + f_{L,4}(x, y, m) \\
\log(\nu_{\sigma^2}) = \beta_{S,0} + f_{S,1}(\mu_T) + f_{S,2}(m) + f_{S,3}(x, y)
\end{cases},
\end{equation}

\noindent where $\mu_{\sigma^2_T}$ and $\nu_{\sigma^2_T}$ indicate the location and scale parameters of the gamma distribution of $\sigma^2_T$, and together they determine the variance of $\sigma^2_T$, indicated as $\mathbb V(\sigma^2_T)$. Functions $f_{L,j}$ and $f_{S,j}$ indicate the $j^\text{th}$ smooth functions for the location and scale parameters, respectively. To ensure the smooth functions of month, $f_{L,2}(m)$ and $f_{S,2}(m)$, joined smoothly between December and January, the terms use cyclic cubic spline bases. The spatial terms used two-dimensional Duchon splines, a generalization of thin-plate regression splines (p. 221 of @wood_generalized_2017). The smoothness parameters were estimated via REstricted Maximum Likelihood [REML$;$ p. 83 of @wood_generalized_2017]. See the Data Availability Statement for additional information on the code used to fit the model.

We simulated hourly variation in future years by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the gamma GAMLS. For each month within each year from 2025 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantiles by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile.

@@

\noindent We estimated percent forest cover and distance from water using the rasters created by @tuanmu_global_2014. We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needle-leaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively). We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package. Finally, we obtained two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [v. 0.99.0, @hollister_elevatr_2023]. We used a raster with a zoom of 6 (a resolution of 0.009$^\circ$) for model fitting and one with a zoom of 3 (a resolution of 0.08$^\circ$) for downloading climate change projections via `climatenaR`. All final rasters and code are available in the GitHub repository associated with this manuscript (see the Data Availability Statement).

@@

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [v. 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated each mammal's instantaneous speed at each observed location by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The 18 animals with insufficiently fine sampling to support movement models with finite speed were for one mountain goat, 15 boreal caribou, and two southern mountain caribou (Table \@ref(tab:data-table)).

Since $\texttt{ctmm}\text{'s}$ movement models assume a single moving state with stochastic but non-zero speed, we corrected data-informed speeds so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub (see the Data Availability Statement). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using a $k$-means algorithm with 2 clusters for each species' distribution of detrended speeds. When the algorithm clearly failed to discriminate between states, we estimated the split point using the inflection points in histograms of the detrended speeds (Fig. B1).

## Estimating the effects of temperature on mammals' movement behaviour

\noindent Ambient temperature is only one of the many drivers of mammalian movement behaviour (Fig. \@ref(fig:dags)). Many species adapt their movement rates (e.g., movement frequency and speed) daily or seasonally in response in factors such as solar time, photoperiod, forage availability, reproductive cycles, and predator avoidance. Similarly, ambient temperature also fluctuates throughout the day and across seasons. Therefore, estimating the effects of temperature on movement rates requires accounting for how mammals' response to temperature changes with time of day and day of year [Fig. \@ref(fig:dags)A$;$ @peron_periodic_2017; @buderman_time-varying_2018; @leclerc_determinants_2021; @xu_plasticity_2021]. Similarly, mammals' selection strength for resources depends on ambient temperature, since higher temperatures can promote a selection for refuge from heat [e.g., thicker forest cover, higher elevation, proximity to water$;$ @attias_effects_2018; @giroux_activity_2023].

The two following sections detail the statistical framework and models we used to estimate the effect on temperature on the six species' movement behavior. To assess the importance of including temperature as an explicit covariate (as ooposed to including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models via analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017. See Appendix B for additional information.

```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature on each species' movement behaviour. (A) Ambient temperature affects mammals' movement rates (i.e. probability of moving, speed when moving, and distance traveled), but temperature is influenced by time of day and day of year. Additionally, the effects of temperature on movement rates depend on time of day and season, since animals may respond to temperatures differently at different times of day and or days of year. Finally, the effect of time of day changes seasonally due to changes in photoperiod. (B) Similarly, habitat selection depends on the availability and selection of resources (forest cover, elevation, and distance from water), but the selection strengh for each resource is conditional on temperature. For example, an animal may avoid forested areas unless it is seeking refuge from extreme temperatures.", fig.height=8, fig.width=20, out.width='\\linewidth'}
library('dplyr')   # for data wrangling
library('ggplot2') # for fancy plots
library('cowplot') # for plots in grids

# movement rates
boxes_a <- tibble(xmin = c(0, 10, 5, 5),
                  xmax = xmin + 4,
                  ymin = c(10, 10, 5, 0),
                  ymax = ymin + 1,
                  fill = c('#BBBBBB80', '#BBBBBB80', '#FF8C00AA', '#FF0000CC'),
                  label = c('Time of day', 'Day of year', 'Movement rates',
                            'Temperature'))

arrows_a <- tibble(id = sort(rep(1:6, 2)),
                   x = c(10, 4, 4, 6, 10, 8, 2, 6, 7, 7, 12, 8),
                   y = c(10.5, 10.5, 10, 6, 10, 6, 10, 1, 1, 5, 10, 1))
